Phenotypic evolution is more constrained in simpler food webs

### 2019-03-29 manuscript.Rmd
edit methods and a bit of results to reduce emphasis on food-web complexity. Also altered wording to make clear that I standardized traits for all treatments.

```{r Conceptual, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:Conceptual}Experimental manipulation of food-web complexity associated with a leaf-galling midge (*Iteomyia salicisverruca*) feeding on the willow *Salix hookeriana*. Black arrows denote the flow of energy in this network of trophic interactions. In the complex food-web treatment, we allowed the full suite of egg and larval parasitoids to impose selection. In the simple food-web treatment, we used mesh bags to exclude the guild of larval parasitoids, only allowing the egg parasitoid (*Platygaster* sp.) to impose selection. Larval parasitoids include the following species (from left to right): *Mesopolobus* sp. (Family: Pteromalidae); *Tetrastichus* sp. (Family: Eulophidae); and *Torymus* sp. (Family: Torymidae)."}
knitr::include_graphics("analyses/complex_simple_foodwebs_v2.jpeg")
```

@@

## Manipulating Food-web Complexity

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. For the complex food-web treatment (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. For the simple food-web treatment, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989]. In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by parasitoids, we excluded unknown sources of mortality. For the food-web treatment that excluded larval parasitoids (simple food web), we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

Our analysis consisted of three parts. First, we used generalized linear mixed models (GLMM) to quantify selection surfaces \textemdash linear and nonlinear relationships between absolute fitness ($W$) and phenotypic traits ($z_i$) of individuals \textemdash in each food-web treatment. Second, we translated selection surfaces into the scale of relative fitness ($w$) in order to calculate selection gradients. Third, we used our estimates of selection gradients to characterize the slope and curvature of the adaptive landscape. Note that inferring adaptive landscapes from selection surfaces assumes that trait distributions are multivariate normal [@Lande1983]. To approximate this assumption, we log-transformed clutch size and square-root transformed oviposition preference. We then scaled all phenotypic traits (mean=0 and SD=1) in order to calculate standardized selection gradients that were comparable across traits and with other studies of natural selection.

@@

**Evolutionary constraints**: We quantified the effects of food-web complexity on evolutionary constraints by inspecting the slope and curvature of the adaptive landscape. The number of selective constraints is determined by the slope of the adaptive landscape, which in our study corresponds to:

$$\text{Slope} = \beta = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} $$
By comparing the number of traits that determine the slope (i.e., nonzero directional selection gradients) in simpler vs. more complex food webs, we can infer the effect of food-web complexity on selective constraints.

@@

Rather than imposing selection, parasitoids may influence the expression of herbivore traits which could bias measurements of selection. In our system, it was plausible that parasitoids may influence chamber diameter by altering larval feeding behavior or killing larvae before they complete their development. To estimate this potential bias, we subset our data to only include galls where there was variation in larval survival within the same gall (i.e. 1 > mean survival > 0). If we assume that larvae within each gall should have similar chamber diameters because they come from the same clutch and experience the same local environment (an assumption our data supports: gall ID explains 54% of the variance in chamber diameter), then the relationship between chamber diameter and larval survival in this data subset represents the effect of parasitism on trait expression (i.e. bias). We used a GLMM with the same structure as described above except that we modelled only a linear relationship between chamber diameter and larval survival ($\alpha_{\text{Diam}}$). We detected a positive bias in both food-web treatments (complex $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebComplex:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.high,2)`]; simple $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebSimple:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.high,2)`]), indicating that unadjusted relationships would overestimate the strength of selection on chamber diameter. To account for this bias, we substracted our mean estimates of bias from our estimates with the full dataset prior to calculating chamber diameter's trait-fitness relationship and directional selection gradient.

@@

Once parasitized, the gall phenotype becomes the extended phenotype of the egg parasitoid. This extended phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the extended phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify trait-fitness relationships and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web complexity on the egg parasitoid's adaptive landscape ---we can only estimate the selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Simpler food webs increase selective constraints

@@

## Simpler food webs increase genetic constraints

@@

To estimate the net effect of food-web complexity on the curvature ($C$) of the adaptive landscape, we only retained nonzero selection gradients (i.e. 95\% CI did not overlap zero) prior to calculating the curvature ($\gamma - \beta \beta^\text{T}$) in each food-web treatment. We found that the curvatures of the adaptive landscape exhibited the following structures:

@@

Only gall midge preference influenced the probability of observing egg parasitoids in complex vs. simple food webs (table \ref{Table:ExtendedGradients}). Specifically, the impact of larval parasitoids increased nonlinearly with higher gall midge densities (fig. \ref{fig:EggPtoid_Selection}).

>>>>

```{r Conceptual, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:Conceptual}Experimental manipulation of food-web structure associated with a leaf-galling midge (*Iteomyia salicisverruca*) feeding on the willow *Salix hookeriana*. Black arrows denote the flow of energy in this network of trophic interactions. In the complex food-web treatment, we allowed the full suite of egg and larval parasitoids to impose selection. In the simple food-web treatment, we used mesh bags to exclude the guild of larval parasitoids, only allowing the egg parasitoid (*Platygaster* sp.) to impose selection. Larval parasitoids include the following species (from left to right): *Mesopolobus* sp. (Family: Pteromalidae); *Tetrastichus* sp. (Family: Eulophidae); and *Torymus* sp. (Family: Torymidae)."}
knitr::include_graphics("analyses/complex_simple_foodwebs_v2.jpeg")
```

@@

## Manipulating Food-web Structure

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. As a control (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. To exclude the guild of larval parasitoids, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989]. In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by parasitoids, we excluded unknown sources of mortality. For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

Inferring adaptive landscapes assumes that trait distributions are multivariate normal [@Lande1983]. 
To approximate this assumption, we log-transformed clutch size and square-root transformed oviposition preference. To ensure that our estimates of selection were comparable across traits and with other studies, we scaled all phenotypic traits (mean=0 and SD=1) prior to our analyses (detailed below).

Our analysis consisted of three parts. First, we used generalized linear mixed models (GLMM) to quantify selection surfaces \textemdash linear and nonlinear relationships between absolute fitness ($W$) and standardized phenotypic traits ($z_i$) of individuals \textemdash in each food-web treatment. Second, we translated selection surfaces into the scale of relative fitness ($w$) in order to calculate standardized selection gradients. Third, we used our estimates of selection gradients to characterize the slope and curvature of the adaptive landscape.

@@

**Evolutionary constraints**: We quantified the effects of food-web structure on evolutionary constraints by inspecting the slope and curvature of the adaptive landscape. The number of selective constraints is determined by the slope of the adaptive landscape, which in our study corresponds to:

$$\text{Slope} = \beta = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} $$
By comparing the number of traits that determine the slope (i.e., nonzero directional selection gradients) in our food-web treatments, we can infer the effect of food-web structure on selective constraints.

@@

Rather than imposing selection, parasitoids may influence the expression of herbivore traits which could bias measurements of selection. In our system, it was plausible that parasitoids may influence chamber diameter by altering larval feeding behavior or killing larvae before they complete their development. To estimate this potential bias, we subset our data to only include galls where there was variation in larval survival within the same gall (i.e. 1 > mean survival > 0). If we assume that larvae within each gall should have similar chamber diameters because they come from the same clutch and experience the same local environment (an assumption our data supports: gall ID explains 54% of the variance in chamber diameter), then the relationship between chamber diameter and larval survival in this data subset represents the effect of parasitism on trait expression (i.e. bias). We used a GLMM with the same structure as described above except that we modeled only a linear relationship between chamber diameter and larval survival ($\alpha_{\text{Diam}}$). We detected a positive bias in both food-web treatments (complex $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebComplex:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.high,2)`]; simple $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebSimple:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.high,2)`]), indicating that unadjusted relationships would overestimate the strength of selection on chamber diameter. To account for this bias, we substracted our mean estimates of bias from our estimates with the full dataset prior to calculating chamber diameter's trait-fitness relationship and directional selection gradient.

@@

Once parasitized, the gall phenotype becomes the extended phenotype of the egg parasitoid. This extended phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the extended phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify trait-fitness relationships and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate the selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Loss of species interactions increase selective constraints

@@

## Loss of species interactions increase genetic constraints

@@

To estimate the net effect of food-web structure on the curvature ($C$) of the adaptive landscape, we only retained nonzero selection gradients (i.e. 95\% CI did not overlap zero) prior to calculating the curvature ($\gamma - \beta \beta^\text{T}$) in each food-web treatment. We found that the curvatures of the adaptive landscape exhibited the following structures:

@@

The exctinction of larval parasitoids only altered the relationship between gall midge preference and the probability of observing egg parasitoids (table \ref{Table:ExtendedGradients}). Specifically, the impact of larval parasitoids increased nonlinearly with higher gall midge densities (fig. \ref{fig:EggPtoid_Selection}).
